\newenvironment{effectize}{Effect: \begin{itemize}}{\end{itemize}}
\newcommand{\effect}{\item[]}
\newenvironment{effectblock}{\begin{itemize}}{\end{itemize}}

\newcommand{\regeffect}[1]{\effect $R_r \leftarrow #1$}
\newcommand{\regeffects}[1]{\begin{effectize}\regeffect{#1}\end{effectize}}

\newcommand{\bitregeffect}[1]{\effect $R_{r,i} \leftarrow #1$}
\newcommand{\bitregeffects}[1]{\begin{effectize}\bitregeffect{#1}\end{effectize}}

\newcommand{\priveffect}{\effect if $U_C = 1$ then trigger a \name{Privileged Instruction Fault}}

\newcommand{\rrrformat}[1]{
Format:
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Bits & 31..26 & 25..21 & 20..16 & 15..11 & 10..0\\
\hline
Value & #1 & x & y & r & (ignored)\\
\hline
\end{tabular}
}

\newcommand{\rriformat}[1]{
Format:
\begin{tabular}{|c|c|c|c|c|}
\hline
Bits & 31..26 & 25..21 & 20..16 & 15..0\\
\hline
Value & #1 & x & r & y\\
\hline
\end{tabular}
}

\newcommand{\jformat}[1]{
Format:
\begin{tabular}{|c|c|c|}
\hline
Bits & 31..26 & 25..0\\
\hline
Value & #1 & offset\\
\hline
\end{tabular}
}

\newcommand{\jrformat}[1]{
Format:
\begin{tabular}{|c|c|c|c|}
\hline
Bits & 31..26 & 25..21 & 20..0\\
\hline
Value & #1 & dest & (ignored)\\
\hline
\end{tabular}
}

\newcommand{\brformat}[1]{
Format:
\begin{tabular}{|c|c|c|c|c|}
\hline
Bits & 31..26 & 25..21 & 20..16 & 15..0\\
\hline
Value & #1 & x & y & offset\\
\hline
\end{tabular}
}

\newcommand{\ldstformat}[1]{
Format:
\begin{tabular}{|c|c|c|c|c|}
\hline
Bits & 31..26 & 25..21 & 20..16 & 15..0\\
\hline
Value & #1 & x & r & y\\
\hline
\end{tabular}
}

\newcommand{\noargformat}[1]{
Format:
\begin{tabular}{|c|c|c|}
\hline
Bits & 31..26 & 25..0\\
\hline
Value & #1 & (ignored)\\
\hline
\end{tabular}
}

\newcommand{\mvspformat}[1]{
Format:
\begin{tabular}{|c|c|c|c|c|}
\hline
Bits & 31..26 & 25..21 & 20..16 & 15..0\\
\hline
Value & #1 & (ignored) & r & z\\
\hline
\end{tabular}
}

\chapter{Instruction Set}

The instructions of the \eco operate directly on the functional components described in the previous chapter. They can be subdivided into groups of instructions that work in a similar way:
\begin{itemize}
\item {\it Computation}: These instructions compute a function of values stored in general-purpose registers or encoded directly into the instruction and store the result in a general-purpose register.
\item {\it Control Flow}: These instructions affect the \pc in various ways.
\item {\it Load/Store}: These instructions transfer data from or to RAM locations or peripheral device registers.
\item {\it System}: Special instructions for \pswx, \mmux, or exception operation.
\end{itemize}

\section{Definitions}

Some definitions are useful when explaining the effect of an instruction: An \definition{immediate} value is a value encoded directly into the instruction. A \definition{register value} is a 32-bit value taken from a general-purpose register. The interpretation of such values is up to the instruction.

A register value is referred to by an instruction by an immediate value that denotes the register number. If $x$ is a 5-bit immediate value, then $R_x$ shall denote the corresponding register value, and $R_x \leftarrow ...$ shall denote an assignment to this register. Similarly, $S_i$ denotes special purpose register \#i. $R_{i,j}$ and $S_{i,j}$ denote specific bits of a register. As a special rule, an assignment to $R_0$ has no effect since that register is not writeable.

\section{General Execution Loop}

The \eco executes the following loop to perform its task:
\begin{itemize}
\item Remember the current value of the \pc register. If any exception occurs before the instruction is finished, this value is placed in register \#30 such that the current instruction can be restarted.
\item Load the current instruction from the virtual address stored in the \pcx. If that address is not word-aligned, then an \name{Invalid Address Exception} occurs. Otherwise, if it is a privileged address and the CPU is in user mode, then a \name{Privileged Address Exception} occurs. Otherwise, it is mapped to a physical address by the \mmux, which may trigger a \name{TLB Miss Exception} or a \name{TLB Invalid Exception}. All these exceptions cause the faulting \pc value to be stored in the \name{TLB Bad Address Register}. Note that a \name{TLB Write Exception} cannot occur since the instruction fetch is a read access.
\item Increase the \pc by 4.
\item If the opcode in the instruction word does not denote a valid instruction, then an \name{Illegal Instruction Fault} is triggered.
\item Decode and execute the instruction. Any fault triggered during this step immediately stops execution of the current instruction and transfers control to the fault service routine.
\item Remember the new value of the \pc register. If any interrupt occurs in the next step, this value is placed in register \#30 such that control can return to the next instruction.
\item Test for interrupts. If an interrupt is signalled and admitted (\myref{2}{ien}), then control is transferred to the service routine (\myref{2}{accept_exception}).
\end{itemize}

\input{isa/comp}
\input{isa/jump}
\input{isa/ldst}
\input{isa/system}
