\begin{thebibliography}{sotief}

\bibitem{tanenbaum}
Andrew S. Tanenbaum, Albert S. Woodhull. 
{\it Operating Systems: Design and Implementation}.
second edition, Prentice-Hall, 1997

\bibitem{lions}
John Lions.
{\it Lions' Commentary on UNIX 6th Edition with Source Code}.
Peer-to-Peer Communications, 1996

\bibitem{bach}
Maurice J. Bach.
{\it The Design of the UNIX Operating System}.
Prentice-Hall, 1986

\bibitem{c_programming_language}
Brian W. Kernighan, Dennis M. Ritchie.
{\it The C Programming Language}.
second edition, Prentice-Hall, 1988

\bibitem{retargetable_c_compiler}
Christopher W. Fraser, David R. Hanson.
{\it A Retargetable C Compiler: Design and Implementation}.
Addison-Wesley, 1995

\bibitem{standard_c_library}
P. J. Plauger.
{\it The Standard C Library}.
Prentice Hall P T R, 1992

\bibitem{see_mips_run}
Dominic Sweetman.
{\it See MIPS Run}.
Morgan Kaufmann, 1999

\bibitem{mips_risc_architecture}
Gerry Kane, Joe Heinrich.
{\it MIPS RISC Architecture}.
Prentice-Hall, 1992

\bibitem{dlx}
Philip M. Sailer, David R. Kaeli.
{\it The DLX Instruction Set Architecture Handbook}.
Morgan Kaufmann, 1996

\bibitem{patterson_henessy}
David A. Patterson, John L. Hennessy.
{\it Computer Organization \& Design: The Hardware/Software Interface}.
second edition, Morgan Kaufmann, 1998

\bibitem{henessy_patterson}
John L. Hennessy, David A. Patterson.
{\it Computer Architecture: A Quantitative Approach}.
second edition, Morgan Kaufmann, 1996

\bibitem{ashenden}
Peter J. Ashenden.
{\it The Designer's Guide to VHDL}.
Morgan Kaufmann, 1996

\bibitem{moorby}
Donald E. Thomas, Philip R. Moorby.
{\it The Verilog Hardware Description Language}.
fifth edition, Kluwer Academic Publishers, 2002

\end{thebibliography}
